---
title: "Chapter 3: Exploring the Historical Data"
author: "Dan Nguyen"
date: "August 17, 2015"
output:
  md_document:
    variant: markdown_github
---

## Setup

```{r, message = FALSE}
# Load libraries
library(ggplot2)
require(grid)
library(dplyr)
library(lubridate)
library(rgdal)

# load my themes:
source("./myslimthemes.R")
theme_set(theme_dan())

# create a data directory
dir.create('./data')

```

Download the map data as before:

```{r}
fname <- "./data/cb_2014_us_state_20m.zip"
if (!file.exists(fname)){
  url <- "http://www2.census.gov/geo/tiger/GENZ2014/shp/cb_2014_us_state_20m.zip"
  print(paste("Downloading: ", url))
  download.file(url, fname)
}
unzip(fname, exdir = "./data/shp")
```

Load the map data and make two versions of it:
- `states_map` removes all non-U.S. states, such as Washington D.C. and Guam:
- `cg_map` removes all non-contiguous states, e.g. Alaska and Hawaii.

```{r}
us_map <- readOGR("./data/shp/cb_2014_us_state_20m.shp", "cb_2014_us_state_20m")
states_map <- us_map[!us_map$STUSPS %in%
                        c('AS', 'DC', 'GU', 'MP', 'PR', 'VI'),]
cg_map <- states_map[!states_map$STUSPS %in% c('AK', 'HI'), ]
```


### Download the historical quake data

TK explanation.


```{r}
fn <- './data/usgs-quakes-dump.csv'
zname <- paste(fn, 'zip', sep = '.')
if (!file.exists(zname) || file.size(zname) < 2048){
  url <- paste("https://github.com/dannguyen/ok-earthquakes-RNotebook",
    "raw/master/data", zname, sep = '/')
  print(paste("Downloading: ", url))
  # note: if you have problems downloading from https, you might need to include
  # RCurl
  download.file(url, zname, method = "libcurl")
}
unzip(zname, exdir="data")
# read the data into a dataframe
usgs_data <- read.csv(fn, stringsAsFactors=FALSE)
```


```{r, message = F}
# Remove all non earthquakes and events with magnitude less than 3.0:
quakes <- usgs_data %>% filter(mag >= 3.0, type == 'earthquake')
# Use lubridate to add a `year` and era column
quakes <- quakes %>% mutate(year = year(time)) %>%
    mutate(era = ifelse(year <= 2000, "1995-2000",
      ifelse(year <= 2005, "2001-2005",
      ifelse(year <= 2010, "2006-2010", "2011-2015"))))

# Create a spatial data frame----------------------------------
sp_quakes <- SpatialPointsDataFrame(data = quakes,
                          coords = quakes[,c("longitude", "latitude")])
sp_quakes@proj4string <- states_map@proj4string

# subset for earthquakes in the U.S.
xdf <- over(sp_quakes, states_map[, 'STUSPS'])
states_quakes <- cbind(sp_quakes, xdf) %>% filter(!is.na(STUSPS))
```


Plot a histogram of all of the earthquakes by state since 1995, and highlight Oklahoma in orange:

```{r, message = F}
ggplot(states_quakes, aes(STUSPS)) + geom_histogram(binwidth = 1, width = 0.5) +
  geom_histogram(data=filter(states_quakes, STUSPS == "OK"), binwidth = 1, fill="orange", width = 0.5) +
   theme(axis.text.x = element_text(angle = 70)) + ggtitle("U.S. Earthquakes Since 1995")
```

Reduce the noise by filtering out states. Only a few states have a significant number of earthquakes.

Order the x-axis by count: http://stackoverflow.com/questions/20041136/how-to-avoid-ggplot-sorting-the-x-axis-while-plotting-geom-bar



```{r, message = F}
qcounts <- states_quakes %>% group_by(STUSPS) %>% summarise(count = n()) %>%
    arrange(desc(count))
# create a vector
qcounts$x1 <- factor(qcounts$STUSPS, levels = qcounts$STUSPS)
ggplot(qcounts, aes(x1, count)) + 
  geom_bar(width = 0.5, stat = "identity") +
  geom_bar(data = filter(qcounts, STUSPS == "OK"),stat = "identity", fill = "orange", width = 0.5) +
   theme(axis.text.x = element_text(angle = 70)) + ggtitle("U.S. Earthquakes Since 1995")
```

Pick top 12:

```{r}
top_states <- head(qcounts, 12)$STUSPS
top_quakes <- filter(states_quakes, states_quakes$STUSPS %in% top_states)
```




Plot a histogram of the earthquakes by state since 2010.


```{r, message = F}
top_quakes$x1 <- factor(top_quakes$STUSPS, levels = as.vector(top_states))
top_quakes.ge2010 <- filter(top_quakes, year >= 2010) 
ggplot(top_quakes.ge2010, aes(x1)) + 
  geom_histogram(binwidth = 1) +
  geom_histogram(data=filter(top_quakes.ge2010, STUSPS == "OK"), binwidth = 1, fill="orange") +
  ggtitle("U.S. Earthquakes Since 2010")
```

## Within comparisons: OK year over year

ObvTK there's a lot of action since the 2010s. Let's get a closer look:

```{r}
ggplot(states_quakes
```




## Full comparison: OK, number of earthquakes

## Full/width, OK vs States,






## Something


```{r}

top_states_quakes <- filter(states_quakes, STUSPS %in% top_states)
g <- ggplot(top_states_quakes, aes(x = as.character(year))) +
  geom_histogram(binwidth = 1, fill = "#999999", width = 0.5)  +
  geom_histogram(data=filter(top_states_quakes, year >= 2012), binwidth = 1, fill="#990000", width = 0.5) +
  geom_histogram(data=filter(top_states_quakes, year == 2015), binwidth = 1, fill="#BB9999", width = 0.5)

g + facet_wrap(~ STUSPS, ncol = 4) + theme_dan_grid()
```



## TKTK pie



```{r}



pct_quakes <- states_quakes %>%
  filter(STUSPS %in% top_states) %>%
  group_by(era, STUSPS) %>%
  summarise(count = length(STUSPS)) %>%
  group_by(STUSPS) %>% mutate(pct = count * 100/sum(count))

ggplot(pct_quakes, aes(x = factor(1), y = pct, fill = era)) +
  scale_fill_manual(values = c("#999999", "#666666", "#333333", "red")) +
  geom_bar(width = 1, stat = "identity", size = 0.2 , aes(order = desc(era))) +
  coord_polar(theta = "y") +
  facet_wrap(~ STUSPS, ncol = 4)  + theme_dan_map()
```




<!--
To render this file:
library(rmarkdown)
setwd("~/Dropbox/rprojs/ok-earthquakes-Rnotebook/")

this_file <- 'chapter-3-exploring-historical-data.Rmd'
render(this_file, output_dir = './builds',
  html_document(toc = TRUE, self_contained = F))

render(this_file, output_dir = './builds',
  md_document(variant = "markdown_github",
              preserve_yaml = TRUE))
-->



